Notes on Semi-Supervised `Novelty Detection'

### 2011-04-08 semisupervised/notes.tex
Ideas and a lot of changes on the report.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Threshold Analysis}
\label{sec:threshold-analysis}

\begin{equation}P(x) = \sum_{c}^{} P(x|c) P(c)\end{equation}

\begin{equation}0 \le P(x|c) P(c) \le P(x)\end{equation}
\begin{equation}0 \le P(c) \le \frac{P(x)}{P(x|c)}\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experiments}
\label{sec:experiment}
\begin{itemize}
\item Define a set of features (all extractable from visual or user-tags).
\item Define a dataset of inputs (eg.: indoor environs at LabelMe)
\item Filter those for \emph{kitchen}.
\item Use ML methods to model $P(features | kitchen)$ and $P(features)$.
\item Test results.
\end{itemize}

\subsection{Simple}
\label{sec:simple-experiment}
\note{This first experience is very small and was only used to better explain the effects of the threshold functions.}

For a better understanding of how the knowledge of the unconditional probability $P(x)$ can help to improve the detection threshold
a very small toy experiment was run:

Using a simple distribution, described in \autoref{tab:simple-distribution},
labelled ($known = \{kitchen, corridor\}$)
and unlabelled samples where drawn.
$P(x|class \in known)$ and $P(x)$ where then approximated using simple accumulation tables
(only 6 diferent inputs).
The input space was then sorted according to the presented thresholds obtaining
the results as seen on \autoref{tab:simple-data-sorted}.

\begin{table}
\begin{center}
\input{simple-data/explain}
\end{center}
\caption{\label{tab:simple-distribution}Distribution used on the simple experiment. Each column cell shows $P(feature|class)$}
\end{table}

@@

This simple toy experience demonstrates the importance of using $P(x)$ to estimate a novelty
threshold and it importance when dealing with features that have a tedency to be highly bias
such as the semantic data used on our sample.

\subsection{Dora with world model}

This experience is similar to the simple experiment run in \autoref{sec:simple-experiment}.
But this time using a more realistic distribution (\autoref{tab:dora-world-distribution}) with
11 diferent room categories and with a feature space containing semantic data for:
room size, room shape, visual appearance and 6 diferent objects.
Also, the features are statistically independent from each other given the room category.

The system objective was to detect novel data from a given known set of 6 categories.
In order to use drawn results two tests where run:
\begin{enumerate}
\item Using perfect knowledge of the probability functions both thresholds where compared for performance (\autoref{tab:dora-data-perfect}).
\item A simple model was used to learn the probability functions and the thresholds were compared (\autoref{tab:dora-data-simple}).
\end{enumerate}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{Threshold Analysis}
\label{sec:threshold-analysis}

\subsection{Bounding $P(c)$}
\note{This was just a random though... if we can bound $P(c)$ in some way then we get
      some information on how sure we are about a decision. Though its only possible
      to bound it from above, and even that not sure if its computational feasible.}

\begin{equation}P(x) = \sum_{c}^{} P(x|c) P(c)\end{equation}
\begin{equation}0 \le P(x|c) P(c) \le P(x)\end{equation}
\begin{equation}0 \le P(c) \le \frac{P(x)}{P(x|c)}\end{equation}

\subsection{What does a ROC curve looks like for an optimal detector}
\note{The ROC curve shape shows if a detector is perfect given an input space.
      (Or at least I think so (though I haven't write any proof)).
      I guess somewhere there should be literature reference to such a simple concept}

Proof would be something like:

An element of the input space is responsible for a vector change of $(TP, FP)$ in the ROC
curve.
A threshold functions orders the input space.
And the best is to take the greedy approach of sorting the input by $TP/FP$.
That is equivalent to a convex ROC curve.

This leads to the important property:

whatever threshold function you use,
if you get a convex ROC curve, there is no way to perform better using that input space.

At the same time we can proof an ordering on $TP/FP$ to be the same as $TP/(FP+TP)$.
Which is coherent with the threshold of: $P(x|concept)/P(x)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{Evaluation and Experiments}
\label{sec:experiment}

This section describes the two experiences that were run to evaluate the methods.
\begin{description}
\item[\autoref{sec:simple-experiment}] presents a toy experience that aims at
  showing the importance of using $P(x)$ to estimate a novelty threshold and
  it importance when dealing with features that have a tendency to be highly
  biased such as the semantic data used on our sample.

\item[\autoref{sec:dora-experiment}] runs comparisons of the presented threshold
  on a real-world scenario, where a robot has to identify room categories based
  on obtained semantic data, such as object presence and room size.
  The experiment is run both with a fictitious distribution and with a real world
  dataset.
\end{description}

Other possible experiences are:
\begin{itemize}
\item run this type of classification on datasets such as \emph{LabelMe} databases,
  either by using user-supplied tags as semantic data or by extracting them with some
  image analysis.
\end{itemize}

\subsection{Simple}\label{sec:simple-experiment}
For a better understanding of how the knowledge of the unconditional probability $P(x)$ can help to improve the detection threshold
a very small toy experiment was run:

Using a simple distribution, described in \autoref{tab:simple-distribution},
labelled ($known = \{kitchen, corridor\}$)
and unlabelled samples where drawn.
$P(x|class \in known)$ and $P(x)$ where then approximated using simple accumulation tables
(only 6 different inputs).
The input space was then sorted according to the presented thresholds obtaining
the results as seen on \autoref{tab:simple-data-sorted}.

\begin{table}[hb]
\begin{center}
\input{simple-data/explain}
\end{center}
\caption{\label{tab:simple-distribution}Distribution used on the simple experiment. Each column cell shows $P(feature|class)$}
\end{table}

@@

\subsection{Dora with world model}
\label{sec:dora-experiment}

This experience is similar to the simple experiment run in \autoref{sec:simple-experiment}.
But this time using a more realistic distribution (\autoref{tab:dora-world-distribution}) with
11 different room categories and with a feature space containing semantic data for:
room size, room shape, visual appearance and 6 different objects.
Also, the features are statistically independent from each other given the room category.

The system objective was to detect novel data from a given known set of 6 categories.
Two tests where run:
\begin{enumerate}
\item Using perfect knowledge of the probability functions both thresholds where compared for performance (\autoref{tab:dora-data-perfect}).
\item A simple model was used to learn the probability functions and the thresholds were compared (\autoref{tab:dora-data-simple}).
\end{enumerate}
